\documentstyle[a4,11pt]{article}
\hyphenation{Inter-active}
\newcommand{\REDUCE}{REDUCE}
\begin{document}
\thispagestyle{empty}
\vspace{4cm}
\begin{center} {\Huge \bf REDUCE 3.8} \end{center}
\vspace{.3cm}
\begin{center} {\Huge \bf (Personal System)} \end{center}
\vspace{.5cm}
\begin{center} {\LARGE \bf Installation and Operations Manual} \end{center}
\vspace{.3cm}
\begin{center} {\Large Using CSL Version 6.00 --- April 2004} \end{center}

\vspace{8cm}
\begin{center}
Codemist Ltd \\
Alta, Horsecombe Vale \\
Combe Down \\
Bath BA2 5QR \\
England \\
Tel/FAX: +44 1225 837430 \\
e-mail: reduce@codemist.co.uk
\end{center}

\vspace{1cm} These instructions cover installation of the Personal
\REDUCE\  3.8 on Windows, Linux/GNU, and Macintosh OSX with XDarwin
computers, and a short User's Guide to the CSL-based \REDUCE\  system.
In general \REDUCE\  requires a hard disk drive: executable binaries and
help will
fill up about *** Mbytes, while a full system including machine readable
documentation needs about *** Mbytes. 

\newpage
\tableofcontents
\newpage

With \REDUCE\ 3.8 Codemist are releasing our new user interface which is
(very nearly) uniform across all platforms.  This interface can be run
as a window on machines that support that interface, with menus and a
scrollable output area, or it can be run in non-windows mode in a
terminal window, DOS-box or xterm.  As the systems are now so similar
we are producing this unified document, with small subsections on the
individual systems idiosyncrasies, covering mainly installation.

\section{Contents of the Distribution}

The standard distribution contains:

\begin{enumerate}
\item \REDUCE\  Licence Agreement
\item Installation and Use Instructions
\item \REDUCE\  Registration Form
\item Media (CR-ROM or DAT tape, or as arranged)
\end{enumerate}

The CD format is ISO9660 format, with Rock Ridge extensions. {\em
  ***Check this.***}  As well as CSL/\REDUCE\  it contains the two support
  packages {\bf FOX} and {\bf gnuplot} which are used as components of
  \REDUCE.  The CD should have the directory layout should be as shown
  in the figure~\ref{dir}.

\begin{figure}
\label{dir}
\begin{verbatim}
REDUCE38 _________ doc
(the CD) |
         |________ Fox-extras
         |
         |________ log
         |
         |________ object-files
         |
         |________ r38.doc
         |
         |________ support-packages
         |         |
         |         |___ libedit-2.6.7
         |         |
         |         |___ mingw
         |         |
         |         |___ tetex
         |              |
         |              |___ fonts
         |                   |
         |                   |___ afm
         |                        |
         |                        |___ adobe
         |                             |
         |                             |.....
         |
         |________ tests
         |
         |________ util

\end{verbatim}
\caption{Directory Layout}
\end{figure}

\subsection{Installation}

The first stage is to copy the necessary files to the location you
require.

\subsubsection{Linux, Macintosh OSX and Unix systems}

On unix-like systems this can be done with the installation script {\tt
  install-personal.sh} {\em ***which does not yet exist! Needs rename***}.
  This will prompt for an installation directory, create it if
  necessary, and copy the necessary files to that directory.  There
  are other files on the CD which are for packages which \REDUCE\ uses,
  or should be distributed according to the GPL or LGPL licence. 

\subsubsection{Windows systems}

{\em ***To be written***}
{\em ***I really would like an install program if not installshield***}


\subsubsection{Hand Linking of \REDUCE\ }

It is possible that the file {\tt r38} in the top level directory
will work directly off the CD, but because of all the problems with
differing libraries, compiler versions and licencing issues, we
recommend that you relink \REDUCE, especially on Linux systems.

The necessary object files are in the {\tt object-files}
directory.  The command {\tt relink.sh} in the top directory will
create a system for your machine.  We cannot guarantee that this will
work for all Linux distributions.  If the relink fails you either have
to follow the instructions below, or contact Codemist.

{\em ***To Be Written***}

\section{Use of \REDUCE\ }

You will probably want to change your PATH to add the \REDUCE\  directory
to your path.  Note that there is an assumption that {\tt r38},
{\tt r38.img} and {\tt r38.doc} are kept together in the same
directory.  It is usual to set an environment variable {\tt reduce} to
the directory where {\tt r38.img} is kept.

Normal interactive use of \REDUCE\ is by issuing the command
\begin{verbatim}
	r38
\end{verbatim}
and after a delay for loading the executable and the preserved image
file, the prompt will appear in a new window.  If instead the command 
\begin{verbatim}
	r38 -w
\end{verbatim}
is used the system will run as a command line system in the terminal.
In either case the prompt will be
\begin{verbatim}
	1:
\end{verbatim}

\noindent The system is then as described in the \REDUCE\ 3.8 Manual.
Try typing
\begin{verbatim}
      in "test/alg.tst";
      quit;
\end{verbatim}
to run the standard simple test.  The file {\tt alg.rlg} in the
{\tt test} directory shows what output should be expected.

Some parts of the system are not resident in the initial system, but
they are automatically loaded on demand.  Parts of the library
need to be loaded by the function load\_package or by using the {\tt
Load Module} menu and its sub-menus.  

\begin{verbatim}
	load_package laplace;
\end{verbatim}

\section{Some Simple Tests}

When you have installed \REDUCE\ here are a few quick tests you might
like to try to see what it can do. Many more examples of \REDUCE's
capabilities are included on other distribution disks:

{\small
\begin{verbatim}
    factorial 200;            % display value of the factorial of 200
    (x+y)^4;                  % expand out a simple algebraic formula
    int(x^6*log x, x);        % perform an integration
    z := mat((0,1),(1,1));    % create a matrix, save it in variable 'z'
    z^40;                     % raise matrix z to the 40th power
    solve(a*x^2+b*x+c, x);    % solve a quadratic equation
    on rounded; precision 70; % select long-precision floating point mode
    pi;                       % inspect pi to 70 digits;
    bye;                      % finish session - exit from REDUCE
\end{verbatim}
}
When things do not behave as you expect or you want to know how to
solve some particular problem you need to read the manual!  If you
have installed all of \REDUCE\ you will find a collection of test and
demonstration files in the {\tt tests} folder with extension
{\tt .tst}, and corresponding output with extension {\tt .rlg}.

\section{Help}

There is a HTML-based help tree in {\tt r38.doc} which can be used
with an external browser, or the browser can be started from the {\tt
  Help} menu on the right of the \REDUCE\ window.  There is also the
complete manual available as a PDF file in {\tt r38.pdf} at top level.

See the section on Help below (section~\ref{help}) for more details.

\section{Interrupts}
\label{Interrupts}

CSL-\REDUCE\ is designed so Control-C is an interrupt which causes a
LISP-level error.  So if a calculation is taking too long, or for some
other reason it is wished to abort it, typing Control-C should return
control to the user.  As in LISP all errors may be caught and handled
by the program; it is possible that this will not rescue control.
Repeated use of Control-C may help.  Note that from time to
time \REDUCE\ has to ``garbage collect'',  {\em i.e.\/} reorganise its memory to
make more space.  While doing this (which can take a couple of
seconds) interrupts will be recognised but response to them will be
queued up until the end of garbage collection.

A Control-G is a similar interrupt, but it also generates a backtrace,
so it is possible to see where the calculation is which is being
aborted.


\section{Filenames and CSL}

CSL, the LISP system on which this implementation of \REDUCE\ is based,
runs on a variety of computers using several different filing systems:
it attempts to perform some conversions that improve consistency
across the different systems.  Those who always use simple filenames,
where the names do not include any punctuation characters, do not need
to be aware of this but the details may help when the file-name
mapping leads to confusion.

When a file name is presented to CSL it performs two transformations.
The first set of transformations apply only to an initial segment of
the file name.  If this is of the form:

\begin{center} \$\{{\em word}\} \end{center}

\noindent then it is replaced by the value of the environment variable 
{\em word}.
Similarly for the form \${\em name} where {\em name} is a non-empty
sequence of characters terminated by one of the three characters `.',
`/' or `$\backslash$' or by the end of the file name.  Note that the
form \$\{{\em word}\} is mainly intended for use when the character
following the `\}' is not one of these separation characters.  The
expected use of this mechanism is to allow environment variables to
specify the location of various files that are to be accessed, thus
avoiding the need for absolute path names in source
files.

A file name starting with
\begin{verbatim}
            ~name
\end{verbatim}
has its initial {\tt \~{ }name} segment replaced by a path to the home
directory of a user called {\tt name}.  The special case where the name is
not present indicates the home directory of the current user, with a
fall back to the value in an environment variable {\tt home}, again
with the current directory if that is not set.  As for `\$' the name
following a `\~{ }' is terminated by `.', `/' or `$\backslash$'.

\section{Window Environment}

\subsection{Menus}

There is a number of menus which are displayed at the top of the
screen, some of which have been mentioned before.  This section
explains briefly what they all are for.   
%% The later
%% menus are removed while \REDUCE\  is performing computations, as for
%% example one cannot load modules or change switches except when \REDUCE\ 
%% is expecting commands.

\subsubsection{File menu}

The {\em File} menu has three kinds of item in it.  The first five
items are for user input and information
\begin{itemize}
\item{Read...} allows a \REDUCE\  input file to be selected using a
file selection dialog.  When {\bf OK} is finally chosen characters are
inserted into the buffer to do a \REDUCE\  {\tt in} command.  The line
ends with a semicolon, but it is not terminated with a newline.  The
user can decide whether to use the semicolon or replace it with a
dollar terminator.  Note that the file name is expanded to a full
path, and so this facility can only be used for files whose full path
is less than about 240 characters.

\item{\em Save...} provides a facility to save the current buffer to a
{\tt .log} file.  This file will contain the current viewable text, as can be
seen by scrolling back.  It does not save older material.  See also
the -L command line option.

\item{\em Save Selection...} provides a facility to save the currently
  selected text to a {\tt .log} file.  

\item{\em Print...} allows the current buffer to be sent to a printer.  A
  dialog is started to select which printer, paper size {\em etc.}.

\item{\em Print Selection...} is similar to Print, but for the selected text
  only.

\item{\em Exit} leaves \REDUCE\ which can also be achieved with
  Control-$\backslash$ or typing {\tt bye;} or {\tt quit;} to the
  main window.

\end{itemize}


%% The rest of the menu is for information.  It displays the time used,
%% and the overhead time, together with the number of garbage
%% collections.

\subsubsection{Edit}

The Edit menu gives access to the usual cut/copy/paste.
Using the mouse it is possible to select text, either by dragging
(moving the mouse with the button depressed) or by double-clicking to
select words.  The menu items are
\begin{itemize}
\item{\em Cut} removes the selected text from the buffer into the clipboard;
\item{\em Copy} copies the selected text into the clipboard;
\item{\em Paste} pastes a copy of the clipboard text into \REDUCE.  This
  can also be achieved with Control-V;
\item{\em Reinput} is in effect s copy followed by a paste.  If the
  cursor is other than in the input line the text is inserted at the end.
\item{\em Select All} selects all the text in the buffer
\item{\em Clear} clears the buffer, which can also be done with Control-L.
  Note this will include removing the prompt if \REDUCE\ is waiting
  for input; 
\item{\em Redraw} or Control-R redraws the window;
\item{\em Home}moves to the start of the window;
\item{\em End} moves to the end of the window.
\end{itemize}
Note that the characters {\tt Home} and {\tt End} move to the start
and end of the current input as it is being composed.

\subsubsection{Font}

This menu affects the look of the \REDUCE\ window.

\begin{itemize}
\item{\em Font...} introduces a dialog to change the font used for characters
  in the \REDUCE\ window.  Which fonts are offered will depend on
  which fonts are installed on your computer;
\item{\em Reset Font} restores the font selection to the default we use;
\item{\em Reset Window} restores the size of the window to the defaults..
\end{itemize}


\subsubsection{Break}
The menus for user controlled interrupts.

\begin{itemize}
\item{\em Break}\label{Break} allows a calculation to be
interrupted, as described in section~\ref{Interrupts}.  There may be a
short delay before the interrupt if noticed, but then an interrupt
message will appear in the main window.  This can also be done with
Control-C. 

\item{\em Backtrace} is the same as Break but also produces a
  programmer's backtrace to show where the calculation was stopped.
  Control-G is the same action.

\item{\em Pause} or Control-S stops the output of characters to the
  screen, until a {\em Resume} or Control-Q is made.

\item{\em Resume} or Control-Q reverses the {\em Pause}.

\item{\em Go/Stop} provides a different pausing interface, acting as a
  flip-flop.

\item{\em Discard Output} does not display characters that are
  produced but have not yet reached the screen.  This can be useful in
  cases of a great deal of tracing.  Control-O does the same.
\end{itemize}

\subsubsection{Load Module}
\label{Library}
\label{Load}

The {\em Load Module} menu presents the user with a list of the
modules which may be loaded into \REDUCE\ divided into 7 submenus by
initial letter. 
%Modules which have already been loaded are marks 
%with a check sign.  
Selecting one of the modules causes the string
\begin{verbatim}
        load_package "name";
\end{verbatim}
to be inserted into the input stream for the named package.

A package when loaded will have its name greyed out.

\subsubsection{Switch}

\REDUCE\  has a number of switches.  The ones in the delivered system are made
available with the {\em Switches} menu.  The initial state is
indicated as the switches which are ON are marked with a tick sign.
Selecting a switch inserts characters into the input to change the
sense of the switch.  Note that if the user changes the switches
directly, not via this menu, then the menu will not notice the
changes, and so selecting the switch will set it to the opposite of
what is indicated in the menu, which may not be the same as the
current state.

If a switch is in a module not yet loaded the switch item will be
greyed out.

\subsubsection{REDUCE Update}

For use of this item see section~\ref{update}.

\subsubsection{Help}

The exact format of the submenus depend on the base operating system.
There are two main items, {\em Help} and {\em About}.  On some
Unix-like systems there is also a submenu to allow a browser to be
selected.  On Windows and Macintosh the browser is chosen in the
standard way for those systems.

Note that on Macintosh the browser runs as a Carbon application, and
not as an X one.

The {\em About} menu displays a window with names and copyright
information.

For for details on the use of the Help pages displayed from the {\em
  Help} selection of from typing F1, see section~\ref{help}.

\section{CSL and \REDUCE\ User options}
\label{Options}

Command-line options are not sensitive to the case of the letter
immediately following the `-', and in general if an option can be
written as

\hspace*{1cm} {\large {\bf -X}{\em yyy}} 

\noindent with no space between the -X and its argument yyy, it can
also be written {\large {\bf -X }{\em yyy}} with a space.

The following are recognised.

\vspace{2mm}\noindent {\large \bf -B }
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Run in batch mode, with no interaction or window.
}

\vspace{2mm}\noindent {\large \bf -C }
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{Displays Codemist's copyright notice.
}

\vspace{2mm}
\noindent {\large \bf -Dname=value }
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
As if {\tt (setq name value)} were obeyed just before the start of the
Lisp run.  The characters that form the name are used without any
conversion at all, so variables with unusual spellings can be set up
directly.  It does however mean that it is not possible to predefine a
variable whose name includes an `=' sign.  The value (after the `='
sign) is in effect passed through the regular CSL parser, and so must
obey the regular CSL rules about the formation of valid input.  All
Lisp-level syntax is available, and in particular a variable can be
set to values that are symbols, numbers or strings.  Just -Dname sets
name to T.  It is possible to have many -D (and -U) requests on the
command line, and they are processed from left to right.  Note that
these variables are LISP variables, and not necessarily \REDUCE\  ones.}

\hspace*{.25in}\parbox{4.5in}{Note that extra punctuation may be
needed to prevent your operating system or command line decoder from
doing strange things to command lines that include some characters
(notably `*' and `$\backslash$'); when using Unix you may have to
write -D$\backslash$*xxx=yyy if you want to predefine a variable
called *xxx. The `$\backslash$' escape character in such cases is
interpreted by the Unix shell and not by CSL.
}

\vspace{2mm} \noindent {\large \bf -G(options)}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{Set debugging options.  Currently equivalent to
-D*backtrace.}

\vspace{2mm}
\noindent {\large \bf -I(filename) }
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{Specify a file to scan for fasl files.
The file will be opened read-only.

{\bf -I} can be specified several times, and the files given will
be scanned from left to right when a fasl file is looked for.

The special syntax {\bf -I-} indicates a reference to the ``standard''
image file associated with the current application, which is
r38.exe then this file is r38.img
}

\vspace{2mm} 
\noindent {\large \bf -K(number)}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Specify how much memory CSL should use, in units of Kilobytes.
The number is used to establish the space used for workspace,
and additional space is consumed by the CSL executable image
itself, and perhaps by buffers used for input and output
files, and by the stack.  The number is interpreted in decimal
by default, but can be given in hexadecimal if written
starting with ``0x'' (as in the C language).  The value given
may be subject to rounding (e.g. in some versions of CSL store
is allocated in units of 64K), and if the value given is too
small to allow loading of an image file the value may be
increased without comment.  If {\bf -K} is not given or {\bf -K0} is
specified a default heap size (currently about 4 Mbytes)
will be used, but then this space will be expanded (if
possible) later in the run if store seems to be becoming
full.}

\vspace{2mm}
\noindent {\large \bf -L(file)}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Send transcript of everything sent to the screen to the specified file
as well (as if by doing {\tt lisp spool "file"} at the start, and
hence subsequent use of {\tt lisp spool()} can switch the listing off
or redirect it).}

\vspace{2mm} \noindent {\large \bf -O(filename)}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{See below (section~\ref{OptO})}

\vspace{2mm} \noindent {\large \bf -P}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Profile option.  For some systems this enables a monitoring
system that can gather low level information about where time
is being spent.  For use by expert system tuners only, and
potentially no use to those who do not have a source licence
to CSL.}

\vspace{2mm} \noindent {\large \bf -Q}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Run in quiet mode: no messages at start up, or from fasl
mechanism or garbage collector (unless enabled from within
Lisp). }

\vspace{2mm} \noindent {\large \bf -Rnnn}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Sets the initial random seed, for reproducible runs.  -R0 (the
default) sets the initial seed based on the time of day or some
similar value.} 

\vspace{2mm} \noindent {\large \bf -S}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Sets *plap to true at the start of a run so that compilation
leads to a display of the byte-codes that are generated.
Exactly equivalent to -D*plap. }

\vspace{2mm} \noindent {\large \bf -U(name)}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Unset the variable named.  Sort of inverse to -D(name).}

\vspace{2mm} \noindent {\large \bf -V}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Run in verbose mode.  Messages will be displayed at start-up,
when a fasl file is loaded or when there is a garbage
collection. }

\vspace{2mm} \noindent {\large \bf -W}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
This is used to disable use of the window manager, and instead to run
in a ``glass teletype'' mode.
}

\vspace{2mm}\noindent {\large \bf -X }
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Unset any traps so errors lead to system crashes.  This is used for debugging.
}

\vspace{2mm}\noindent {\large \bf -Y }
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Ignore calls to {\em restart}.
}

\vspace{2mm} \noindent {\large \bf -Z}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
See below: cold-start mode for use by system builders only.}

\vspace{2mm}
\noindent {\large \bf filename}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Make that file available to CSL as its standard input file.
If several files are given the top-level read--eval--print
procedure will be invoked several times, once for each.  Note
that a file being read this way will not be deemed to be a
``terminal'' file (while on some systems the shell redirection
<filename may leave CSL thinking that the input is direct from
a terminal).}

\vspace{2mm} 
\noindent {\large \bf $>$(filename)}
\vspace{1mm}

\hspace*{.25in}\parbox{4.5in}{
Redirects the standard output to the given file.  This option
is present as a fall-back, in that it will usually be the case
that output re-direction can be done by the shell or operating
system, and CSL will not need to worry about it.  This option
will not be enabled in general, on the assumption that
operating system redirection is better.}

\vspace{4mm}

There are limits on the number of files CSL can accept and the number
of symbols it can predefine --- up to 40 source files are permitted, up
to 20 directories in the FASL search path (including the implicit one
present by default at the start of the run), and up to 40 symbols can
be defined or undefined.

\section{Corrections to \REDUCE\ }
\label{update}

A patches file, correcting various problems that have been discovered
since the initial release of \REDUCE\ 3.8, is available for installation in
the Codemist versions of \REDUCE.  To do this you either click the
{\em REDUCE update} menu item, or issue the \REDUCE\ command:
\begin{verbatim}
     update_reduce();
\end{verbatim}
This will install the patches file and make it available in this and
future runs of \REDUCE.  If \REDUCE\ is already up to date it will report this
fact, otherwise it just produces brief messages about contacting the
server where updates live.

If you have any problems with this update procedure, you can download
the relevant file, {\tt patches.fsl}, from {\tt
http://reduce-algebra.com/support/patches} .  This file should be
installed in either your home directory, or the directory where your
\REDUCE\ executable and image files ({\em i.e.\/} {\tt r38.exe} or {\tt
r38} and {\tt r38.img}) live.  The version in your home directory has
precedence if such files exist in both places.  When you next run
\REDUCE, your software is updated with the patches file. If network
failure or other issues cause the {\tt patches.fsl} file to get
corrupted then you can should just delete it and a corrected version
will be fetched the next time you update \REDUCE.

Full details of the issues addressed by patches are available at the web
address given above. The file {\tt patches.txt} describes problems, a file
{\tt patches.tst} is a test script and {\tt patches.rlg} sample output
from the tests. 

If you retrieve these files from {\tt www.reduce-algebra.com} you can just read
{\tt patches.tst} into a freshly-started copy of \REDUCE\ and compare its output
with {\tt patches.rlg}.  The usual \REDUCE\ menus and commands can be used to read
the file and save output from it so that it is convenient to make the
comparison.  When \REDUCE\ is run in a window the FILE/READ menu can be used
to read in the test file, and then FILE/SAVE can save the resulting
output.  If \REDUCE\ is on your path so you can run it from a command prompt
a command such as
\begin{verbatim}
    r38 patches.tst -l patches.log
\end{verbatim}
would read the test file and leave a copy of its output in {\tt patches.log}

\section{Gnuplot}

The following notice is taken from the {\bf gnuplot} distribution.  The
version included here was taken from a generally available software
archive, and up to date versions should be similarly available.  
The distributed version is taken directly from the Internet.  It is not
part of \REDUCE\ , but is supplied as \REDUCE\  can generate output for {\bf
gnuplot} and is the usually graphical output system for \REDUCE\ .

{\small
\begin{verbatim}
/*
 * Copyright (C) 1986 - 1993   Thomas Williams, Colin Kelley
 *
 * Permission to use, copy, and distribute this software and its
 * documentation for any purpose with or without fee is hereby granted,
 * provided that the above copyright notice appear in all copies and
 * that both that copyright notice and this permission notice appear
 * in supporting documentation.
 *
 * Permission to modify the software is granted, but not the right to
 * distribute the modified code.  Modifications are to be distributed
 * as patches to released version.
 *
 * This software is provided "as is" without express or implied warranty.
 *
 * AUTHORS
 *
 *   Original Software:
 *     Thomas Williams,  Colin Kelley.
 *
 *   Gnuplot 2.0 additions:
 *       Russell Lang, Dave Kotz, John Campbell.
 *
 *   Gnuplot 3.0 additions:
 *       Gershon Elber and many others.
 *
 * Send your comments or suggestions to
 *  info-gnuplot@dartmouth.edu.
 * This is a mailing list; to join it send a note to
 *  info-gnuplot-request@dartmouth.edu.
 * Send bug reports to
 *  bug-gnuplot@dartmouth.edu.
 */
\end{verbatim}
}

\subsection{\REDUCE\  Help System}
\label{help}

The \REDUCE\ help is presented in HTML and is viewed via a browser,
either the system default, or as selected from the Reduce Help menu.
The initial page is a list of major topics.
{\em *** would be nice if it had at least a Reduce and/or CSL picture
  and header!!}
There is also a link to an alphabetical index page.


\section{Managing Images}
\label{OptO}

The state of a \REDUCE\ session can be captured through use of the
LISP function {\tt preserve}.  When CSL is first entered it loads a
heap image from the data fork of the application; a call to {\tt lisp
preserve()} dumps the current state of CSL, including all current
variable and function definitions (but not the stack), back onto that
file.  CSL then stops.  Optionally a quoted argument can be given to
{\tt preserve}: this will be treated as the name of a function to be
called when the dumped image is re-loaded, and typically it will be a
replacement for the built-in CSL read--eval--print loop.  In general
for \REDUCE\ this function is {\tt begin}.  If such an argument is not
given, or if {\tt NIL} is specified, then the startup function present
when CSL was loaded is retained.  The InitialImage subfile is kept as
part of the data fork of the application, so those wanting to produce
a modified version of the system will usually need to take a copy of
the entire file.  Alternatively use can be made of the -O option
discussed later and -Z which can start up CSL without the benefit of
any initial Image file.  From within a running system a call to {\tt
restart!-csl} abandons all existing calculations and sets the system
up as if it had just been loaded.  If the argument to {\tt
restart!-csl} is nil then it is as if the -Z option had been given,
and only a very minimal Lisp environment is available.  Otherwise an
initial image is loaded, as at the start of a normal run.  This
function is used in the scripts that rebuild a complete \REDUCE, but
is not expected to be of interest to the casual \REDUCE\ user.

{\em More Needed here???}

\end{document}

[end]
}



