\chapter {Maintaining {\REDUCE}}

{\REDUCE} continues to evolve both in terms of the number of facilities
available, and the power of the individual facilities.  Corrections are
made as bugs are discovered, and awkward features simplified.  In order to
provide users with easy access to such enhancements, a {\em {\REDUCE}
network library\/} has been established from which material can be extracted
by anyone with electronic mail access to the Internet computer network.

In addition to miscellaneous documents, source and utility files, the
library includes a bibliography of papers referencing {\REDUCE} which
contains over 800 entries.  Instructions on using this library are sent to
all registered {\REDUCE} users who provide a network address.  If you
would like a more complete list of the contents of the library, send to
{\em reduce-netlib@rand.org\/} the single line message {\em send index\/} or
{\em help}. The current {\REDUCE} information
package can be obtained from the network library by including on a
separate line {\em send info-package\/} and a demonstration file by
including the line {\em send demonstration}.  If you prefer, hard copies
of the information package and the bibliography are available from the
{\REDUCE} secretary at RAND, 1700 Main Street, P.O. Box 2138, Santa
Monica, CA 90407-2138 ({\em reduce@rand.org}).  Copies of the network
library are also maintained at other addresses.  At the time of writing,
{\em reduce-netlib@can.nl\/} and {\em reduce-netlib@pi.cc.u-tokyo.ac.jp\/}
may also be used instead of {\em reduce-netlib@rand.org}.

A World Wide Web {\REDUCE} server with URL
\begin{verbatim}
	http://www.rrz.uni-koeln.de/REDUCE/
\end{verbatim}
is also supported.  In addition to general information about {\REDUCE}, this
server has pointers to the network library, the demonstration versions,
examples of {\REDUCE} programming, a set of manuals, and the {\REDUCE} online
help system.

Finally, there is a {\REDUCE} electronic forum accessible from the same
networks.  This enables {\REDUCE} users to raise questions and discuss
ideas concerning the use and development of {\REDUCE} with other users.
Additions and changes to the network library and new releases of {\REDUCE}
are also announced in this forum.  Any user with appropriate electronic
mail access is encouraged to register for membership in this forum.  To do
so, send a message requesting inclusion to \\
{\em reduce-forum-request@rand.org}.

