\documentclass [a4paper,11pt]{article}
\title{Packing static tables for compactness and rapid access}
\author{Arthur C Norman}
\begin{document}
\maketitle
\section{Synopsis}
\section{Introduction}
This is a message to record some of the fun I have been having recently, 
in case anybody else ends up interested.

I want to be able to display mathematical formulae nicely. To that end I 
need access to character metrics. In my most recent investigation I am 
expecting you use seven fonts
   Computer Modern Unicode Typewriter Text  (fixed pitch)
   "Odokai" [AR PL New Kai]                 (a range of CJK characters)
   STIX in Regular, Bold, Italic and BoldItalic (a serif face)
   STIX-Math                                (mathematical symbol support)
Between these I am faced with just over 33000 characters. I want tables 
that give character sizes, kern and ligature information and various extra 
stuff relating to the typesetting of mathematics. I would like the tables 
to be compact but fast to access. I am prepared to take the view that I 
will only use these particular fonts - so a user is not entitles to bring 
in another without needing to do substantial revision and rebuilding of 
everything.

By far the biggest table is one one giving character sizes and providing 
indirections into ligature and kern tables. To support fast access I make 
this a hash table. If I hashed each character individually then I would 
have 33000 entries in my table each using space to hold the hash key. For 
the data concerned most characters are arrange in blocks, so I use one 
hash table entry for every run of four characters. This wastes some space 
where the fonts have isolated characters in them, but saves repetition of 
hash keys. I end up with just over 10000 lines active in the hash table, 
so on average each line has three of the four positions in it ii use. It 
happens that using this arrangement also gives me a neat number of bits to 
be used for each hash table entry - 40 bytes which I store as five 64-bit 
words.

\section{Approach}
When establishing a hash table there are typically messy compromises to 
make. One would like a hash function that is cheap and simple to compute 
to save time there, but something too simple may compromise the efficiency 
of the table. One would like as small a table as possible to save space, 
but that will tend to lead to the need for a large number of probes to 
retrieve information. I use a variant of cuckoo hashing, and the main 
point of this note is to record what I view as the amazing way in which 
this lets me meet all my objectives simultaneously.

I introduce three separate hash functions and for any character I insist 
that it end up at one of the locations that these propose for it. Of 
course in unlucky cases two or even all of the functions will yield the 
same value and then the character concerned does not have as much 
flexibility about where it goes. Accepting that as a fact of life, I go 
further. Characters in the typewriter and the mathematical font whose 
codes are in the range U+0000 to U+007f (ie the basic latin alphabet in 
the two fonts I expect to be most heavily used) are treated so that only a 
single hash value is used. Then in the typewriter font I make it such that 
all characters at codepoints up to U+0600 (this covers a serious range of 
European alphabets) and the main range from the Maths font that cover 
European letters and mathematical sumbols only have two distinct hash 
values. Characters expected to arise less commonly - bold, italic, CJK and 
a whole raft of Unicode oddities - are allowed the full three 
possibilities.

The one, two or three locations in the hash table that each character is 
allowed to reside can be thought of as a bipartite graph, and standard 
algorithms (eg Hopcroft-Karp) can seek a maximum matching. If the one that 
is found manages to allocate a location for every single character one has
a hash table where the most important latin characters can be found in one 
probe, essentially all symbols used in mathematics in two (and hance in an 
average of around 1.5), while every character can be looked up in at worst 
three probes (and so in an average of around 2). An unsuccessful search 
for a character not present in any of the fonts only needs three probes. I 
feel that this allows one to believe that performance is good.

The remaining challenge is to ensure that the hash functions used are 
cheap to compute and the table is acceptably compact. I will start with 
the first of these, assuming that the hash table is of size N. The three
hash functions I use are
    h1(k) = k mod N
    h2(k) = k mod N2 + O2  [for some constants N2 and O2]
    h3(k) = (h1(k) + h2(k)) mod N
and if I arrange that N2+O2<=N I can be confident that h2(k) yields a 
value that will lie within table bounds. Especially given that faily often 
only the first one or two of these will need to be computed, and in the 
light of the fact that modern machines compute remainders reasonably fast 
these seem good and simple to implement and not unduly expensive. Observe 
that there is a lot of flexibility in the choice of the constants N2 and 
O2 as well as in choosing the table size N.

Because the font data is static and not expected to change at all often I 
can afford a potentially expensive search to pick N, N2 and O2 so as to 
find a small table. Existing known results about cuckoo hashing indicate 
that for variants like this with three probes one should expect that with 
good has functions it will usually be possible to have a table occupancy 
of around 91%. Here one can worry that the blocky allocation of codes in 
the fonts and the really naive hash functions might make things worst than 
that.

In its most direct form I start with a small table size N and try running 
the maximum matching process for all plausible values of N2 and O2, 
gradually increasing N until I am first able to create a complete table.
When doing this I judge that a tiny value of the secondary modulus N2 is 
not at all liable to be useful since it would not lead to data being 
spread out, so I scan for N2 from 2N/3 up to N-1. I then try all legal 
values of the offset O2 that will avoid h2(k) ever ending up too large. 
This is quite a painfully large search space. To speed it up my code has 
two tricks, both of which introduce risks. The first is that I create 8 
sub-processes and each investigate part of the search space in parallel. I 
avoid heavy duty synchronisation and an effect is that hypothetically one 
part of the search could run ahead and find a packing in a slightly larger 
size hash table before another had complete analysis of a smaller table 
size. Even though that might happen it would probably at worst lead to 
issue of a hash table just one larger than perfection. My other speed-up 
is to use binary search between a tiny table size that is clearly 
impossible to use and a large one where it is easy to fit everything in. 
This obviously speeds things up very usefully, but the binary search 
process really relies on having a monotonic function to investigate if it 
is to guarantee to find the smallest feasible table. Again I have not 
found this to be a big problem in practise. My code is such that a 
complile time option can be used to disable use of parallel search, and an 
alternative entrypoint into the code performs the original naive search 
and so can be used to check solutions found using the optimisations.

\section{Initial results}

With this strategy - and let me remind you that it allows the most 
important characters to be looked up in one probe and almost all the 
relevant rest in only 2 - I find that to store the 10019 keys for my 
characters I end up with a table size of 10057. This is a 99.62% 
occupancy! My metrics table still ends up consuming around 400 Kbytes, but 
I am amazed by how little waste there is.

I have other hash tables to record information about variant and extended 
characters and to help position accents. So for instance a left 
parenthesis need to be associated with four of five characters 
representing gradually larger versions of itself, and then with individual 
characters that can be places adjacent to build up truly huge versions. 
A typical result here is that I have 141 characters for which "extended 
forms" are available, and the hash table set up much as described above 
ends up being of size 143. Again cuckoo hashing with an exhaustive search 
through a range of trivial hash functions has allowed an almost perfect 
hashing scheme to be created.

On a reasonably fast desktop machine running a complete sequential search 
for a good hash table for the 10000 entries I am working with takes of the 
order of a day. The binary search reduces this to a bit under an hour and 
for greater convenience I then seed in and merely verify the best solution 
- and the code that creates the tables then runs in seconds.

A further refinement could be to use not just a simple maximum matching 
algorithm, but one based on weighted graphs so that matchings that 
correspond to placing kets in their first choics position was 
systematically preferred. At present my expectation is that at the high 
table occupancies I am working with the fact that everything fits is 
essentially a fluke based on the exact particular hash parameters in use, 
and that this would not improve things much, but if one backed off to 
lower hash occupancy regimes it might be useful.

My code for all this is in a file "charmetrics.c" and when compiled with a 
"CREATE" flag it can be run to create "charmetrics.h" which contains the 
tables for use with C or C++ and also "charmetrics.red" so that metrics 
could be used directly from Reduce. All these in the csl/cslbase directory 
of the Reduce source tree at sourceforge.

\section{Refinements}

Parallel search using work-farm model

Use linear strong component decomposition before Hopcroft Karp

Strong component with more in A then B must fail

Strong component with only 1 or 2 in B must succeed

If at any stage more items remain in A than in B then failure is certain.

Keep largest component until end in case an earlier one proves impossibility.

Adapt Hungarian code to use spare data representation


\section{Final results}
\section{Conclusions}

\section{References}

\end{document}
