\section{Finding rational or integer zeros of an univariate polynominal}

\iffalse
% \begin{abstract}
  The operators \f{r\_solve} and \f{i_\solve}  compute
  respectively the exact rational or integer zeros of a single
  univariate polynomial using fast modular methods.
%\end{abstract}


  \subsection{Introduction}
  \fi
\ttindex{I\_SOLVE}\ttindex{R\_SOLVE}
\hypertarget{operator:I_SOLVE}{}
\hypertarget{operator:R_SOLVE}{}
The operators \f{r\_solve} and \f{i\_solve}  compute the exact rational zeros
of a single univariate polynomial using fast modular methods.  The
algorithm used is that described by R. Loos (\cite{Loos:1983a}).
The operator \f{r\_solve} computes
all rational zeros whereas the operator \f{i\_solve} computes only
integer zeros in a way that is slightly more efficient than extracting
them from the rational zeros.  The \f{r\_solve} and \f{i\_solve}
interfaces are almost identical, and are intended to be completely
compatible with that of the general \f{r\_solve} operator, although
\f{r\_solve} and \f{i\_solve} give more convenient output when
only rational or integer zeros respectively are required.  The current
implementation appears to be faster than \verb|solve| by a factor that
depends on the example, but is typically up to about 2.

Extension to compute Gaussian integer and rational
zeros and zeros of polynomial systems is planned.


\subsection{The user interface}

The first argument is required and must simplify to either a
univariate polynomial expression or equation with integer, rational or
rounded coefficients.  Symbolic coefficients are not allowed (and
currently complex coefficients are not allowed either.)  The argument
is simplified to a quotient of integer polynomials and the denominator
is silently ignored.

\hypertarget{operator:ARBRAT}{\texttt{ARBRAT}}
Subsequent arguments are optional.  If the polynomial variable is to
be specified then it must be the first optional argument, and if the
first optional argument is not a valid option (see below) then it is
(mis-)interpreted as the polynomial variable.  However, since the
variable in a non-constant univariate polynomial can be deduced from
the polynomial it is unnecessary to specify it separately, except in
the degenerate case that the first argument simplifies to either 0 or
$0 = 0$.  In this case the result is returned by \f{i\_solve} in
terms of the operator \f{arbint} and by \f{r\_solve} in terms of
the (new) analogous operator \f{arbrat}.  The operator
\f{i\_solve} will generally run slightly faster than \f{r\_solve}.
  
The (rational or integer) zeros of the first argument are returned as
a list and the default output format is the same as that used by
\f{solve}.  Each distinct zero is returned in the form of an
equation with the variable on the left and the multiplicities of the
zeros are assigned to the variable \var{root\_multiplicities} as a
list.  However, if the switch \sw{multiplicities} is turned on then
each zero is explicitly included in the solution list the appropriate
number of times (and \var{root\_multiplicities} has no value).
  
\begin{sloppypar}
Optional keyword arguments acting as local switches allow other output
formats.  They have the following meanings:
\begin{description}
\item[\texttt{separate}:] assign the multiplicity list to the global
  variable \texttt{root\_multiplicities} (the default);
\item[\texttt{expand} or \texttt{multiplicities}:] expand the solution
  list to include multiple zeros multiple times (the default if the
  \texttt|multiplicities| switch is on);
\item[\texttt{together}:] return each solution as a list whose second
  element is the multiplicity;
\item[\texttt{nomul}:] do not compute multiplicities (thereby saving
  some time);
\item[\texttt{noeqs}:] do not return univariate zeros as equations but
  just as values.
\end{description}
\end{sloppypar}


\subsection{Examples}

\begin{verbatim}
r_solve((9x^2 - 16)*(x^2 - 9), x);
\end{verbatim}
\[
  \left\{x=\frac{-4}{3},x=3,x=-3,x=\frac{4}{3}\right\}
\]
\begin{verbatim}
i_solve((9x^2 - 16)*(x^2 - 9), x);
\end{verbatim}
\[
  \{x=3,x=-3\}
\]
See the test/demonstration file \verb|rsolve.tst| for more examples.


\subsection{Tracing}

\hypertarget{switch:TRSOLVE}{}
The switch \sw{trsolve}\ttindexswitch[RSOLVE]{TRSOLVE} turns on tracing of the algorithm.  It is off
by default.

