\documentclass[a4paper,11pt]{article}
\title{REDUCE 3.7 Installation and Rebuilding}
\author{Codemist Ltd}
\begin{document}
\maketitle
\section{Introduction}
This document explains how to install, build or rebuild REDUCE 3.7
using the CSL Lisp system.  It covers the cases where your computer
runs Microsoft Windows 95, 98 or NT, or Linux or various brands of
Unix.  Much of the description here will only be relevant to those
who have obtained the ``professional'' version of REDUCE, which comes
complete will all its source code. If only the ``personal'' version is
available all material here that relates to source code may be ignored.

\section{Installation}
\subsection{Windows}
REDUCE is installed by running the {\ttfamily setup} program that forms
part of the distribution. This prompts to discover which components of
REDUCE you want to install and where you would like the files placed.

Note that under Windows two versions of the REDUCE executable are provided.
The one called {\ttfamily r37.exe} runs in a window in the usual way. The
one called {\ttfamily r37c.exe} runs as an old-fashioned command-line
program. This version may prove more useful if REDUCE is to be run from
a script.  If you often use a command-window it may be useful to create
a simple one-line batch file called {\ttfamily r37.bat} and contents
\begin{verbatim}
    x:\r37\lisp\csl\win32\r37 %*
\end{verbatim}
\noindent (where you should replace {\ttfamily x:} with the path to where
the REDUCE files were installed) and place this file in a directory
that is on your search path. You can the launch REDUCE by just issuing
the command {\ttfamily r37}. As an alternative you could add the REDUCE
executable directory to your path. These steps have not been automated
as part of the installatiom procedure since many users will be content
launching REDUCE by clicking on the relevant icon.

\subsection{Unix and Linux}
REDUCE is supplied as a collection of compressed archives. They are called
\begin{tabular}{ll}
{\ttfamily r37exec.tar.gz}   & REDUCE executables plus a few support files \\
{\ttfamily r37xmpl.tar.gz}   & Example files nd matching logs \\
{\ttfamily r37doc.tar.gz}    & Documentation \\
{\ttfamily r37src.tar.gz}    & Source code of REDUCE \\
{\ttfamily r37lisp.tar.gz}   & Source code of the CSL Lisp system \\
\end{tabular}

You will first need to access the files on the CDROM. This may involve
mounting it: you will use whatever procedure you usually do for accessing
a CDROM. I will suppose that this makes the files available with paths
such as \verb+/cdrom/r37exec.tar.gz+ but the exact path will depend on how
your Linux is configured.

Select a directory within which you wish to install REDUCE. Select
it is the current directory. Unpacking the tar files will create a
sub-directory called ``r37'' and everything that is unpacked will be 
within this directory. If you are the system manager you might reasonably 
select {\ttfamily /usr/local}, and if you are an individual user you might
just start in your home directory.  In each case unpack such of the 
archives as you will want using a command like:
\begin{verbatim}
   cd /usr/local
   tar xvfz /cdrom/r37exec.tar.gz
\end{verbatim}

It is suggested that to start with you unpack the executables, the
examples and the documentation.   Between them these will use up between 25
and 30 Mbytes of disc space.  The directory structure created will be
as shown below. Names in bold fixed-pitch type are represent directories,
while those in italic are files.
\[
 \mbox{\ttfamily \bfseries r37} \left[ \begin{array}{l}
      \mbox{\ttfamily \bfseries doc} \left[ \begin{array}{l}
           \mbox{\ttfamily \bfseries html} \left[ \begin{array}{l}
                \mbox{\ttfamily \bfseries \ldots} \\
                \mbox{\em r37.html}
                \end{array} \right. \\
           \mbox{\em install.pdf} \\
           \mbox{\em r37.pdf} \\
           \mbox{\em uguide.pdf} \\
           \mbox{\ttfamily \bfseries util}
           \end{array} \right. \\
      \mbox{\ttfamily \bfseries gnuplot} \\
      \mbox{\ttfamily \bfseries lisp}  \left[ \begin{array}{l}
           \mbox{\ttfamily \bfseries csl}  \left[ \begin{array}{l}
                \mbox{\ttfamily \bfseries csl-c} \\
                \mbox{\ttfamily \bfseries cslbase} \\
                \mbox{\ttfamily \bfseries linux(etc)} \left[ \begin{array}{l}
                     \mbox{\ttfamily \bfseries log} \\
                     \mbox{\em Makefile} \\
                     \mbox{\em r37} \\
                     \mbox{\em r37.img} 
                     \end{array} \right. \\
                \mbox{\ttfamily \bfseries util} \left[ \begin{array}{l}
                     \mbox{\em setexec}
                     \end{array} \right.
                \end{array} \right. \\      
           \end{array} \right. \\
      \mbox{\ttfamily \bfseries log} \\
      \mbox{\ttfamily \bfseries packages} \left[ \begin{array}{l}
           \mbox{\ttfamily \bfseries \ldots} \\
           \mbox{\ttfamily \bfseries support} \left[ \begin{array}{l}
                \mbox{\em patches.red}
                \end{array} \right. \\
           \mbox{\ttfamily \bfseries \ldots}
           \end{array} \right. \\
      \end{array} \right.
\]

If you have unpacked the REDUCE executables you will find a file
{\ttfamily r37/lisp/csl/linux/r37}, and to make it easier to use the system
you might like to install a link to it from some convenient place, eg
using a command similar to one of the two following ones:
\begin{verbatim}
   ln -s /usr/local/r37/lisp/csl/linux/r37 /usr/local/bin/r37
   ln -s ~/r37/lisp/csl/linux/r37 ~/bin/r37
\end{verbatim}
For these to make sense either \verb+/usr/local/bin+ or the \verb+bin+
sub-directory of your home directory should be on your usual search path.
If you set up the a symbolic link as shown then reduce should be able to find
the \verb+r37.img+ file it needs without you having to do anything more.

If you are running on some version of Unix it may be that file-permissions
were not set up when the REDUCE files moved to your machine. Select the
\verb+r37/lisp/csl/util+ directory as current and run the script ``setexec''
\begin{verbatim}
   cd r37/lisp/csl/util
   source setexec
\end{verbatim}
\noindent to correct this if you need to (and you can tell that if you get
a complaint when you try to execute one of the scripts in the \verb+util+
directory).


Documentation is placed in \verb+r37/doc+, with the manual in a form
suitable for use with Adobe's Acrobat reader as \verb+r37/doc/r37.pdf+ or
some more browsable HTML help accessed starting from
\verb+r37/doc/html/r37.html+.

For Linux the provisions of the GNU Public Library License mean that
you have to be provided with object files for all the executables that
together form REDUCE. The reason for this is that you may want to (or
indeed need to) re-link them with a newer version of the Linux system
libraries. Because different Linux installations may have been installed
with different generations of these libraries and some versions are
not compatible with others it {\em could} be that you will find that when
first unpacked the r37 executable will not load and run at all. To
re-build it from object code you should use the script called \verb+relink+
that is in the \verb+r37/lisp/csl/util+ directory:
\begin{verbatim}
   cd r37/lisp/csl/linux
   sh ../util/relink
\end{verbatim}
\noindent To use this script you will have to have a set of Linux development
tools available, specifically \verb.gcc. and the usual C libraries.

If you intend to use the REDUCE interface to {\ttfamily gnuplot} you should
ensure that that package is installed. Gnuplot itself is not part of REDUCE
and it may be easiest for you to fetch and install it in some quite
independent way --- for instance many Linux CDROMS or web mirrors provide it
as a standard option for your installation. The directory \verb+r37/gnuplot+
contains copies of {\ttfamily gnuplot} files as they would be found on
a typical software archive, and if necessary you can unpack and install from
there. For further information about {\ttfamily gnuplot}, its installation and
use, you should check the documentation files that accompany it and the
web sites that they reference.


\section{Testing an installation}
On first installing REDUCE it may make sense to run all the REDUCE test
scripts that have been provided. This should make it possible to verify that
that installation was correct, and it also as a side-effect produces a
log file that compares the speed REDUCE has on your machine with that
observed on a reference one at Codemist. For the initial release of REDUCE 3.7
this reference system is based on an Intel Pentium II running at 400 MHz, and
the tests were run under Windows NT 4.0.

To run the complete tests you need to select the correct part of the REDUCE
tree as your current directory, and in the case of Unix systems you need to
make sure that the executable status of various files are properly set. The
\verb+setexec+ script in \verb+r37/lisp/csl/util+ arranges this. Then
\verb+util/testall+ runs all the tests and \verb+util/checkall+ compares
the results with a set of reference logs. At the end you will find your
own logs from individual packages in \verb+r37/log+, a file showing any
differences between your results and the reference set in
\verb+r37/csl/lisp/<system>/log/checkall.log+ and a summary of timing in
\verb+r37/csl/lisp/<system>/log/times.log+.
\begin{center}
\begin{tabular}{|c|c|} \hline
Windows & Unix \\ \hline
\begin{minipage}{2.3in}
\begin{verbatim}
cd r37\lisp\csl\win32
..\util\testall
..\util\checkall
\end{verbatim}
\end{minipage} & \begin{minipage}{2.3in}
\begin{verbatim}
cd r37/lisp/csl/util
sh ./setexec
cd ../linux
../util/testall
../util/checkall
\end{verbatim}
\end{minipage} \\ \hline
\end{tabular}
\end{center}

On the reference computer running the full set of tests takes between 20
and 30 minutes on an otherwise unused machine. If you interrupt the tests
part way through the script \verb+util/testrest+ (used just as
\verb+util/testall+ is) will continue running tests from wherever you
broke off.  Some workstations may be substantially slower (certainly by up to a
factor of ten) than the reference machine so you may need some patience here.

The \verb+log/checkall.log+ file will contain a section for each test that
was run. The differences reported will certainly include every line that
reports how long anything took.  If you have installed any patches in your
version of REDUCE those too might cause changes in the test output (and these
changes will not then represent errors). Also on different platforms the
exact results from some numeric calculations done in machine-arithmetic will
differ.

The file \verb+log/times.log+  has a line in it for each of the test files,
and this records the time that this test tool on both the local and the
reference computer. It ends up with giving a speed ratio between the
two machines based on all these results.  Each test has two times
associated with it. One includes and the other excludes ``garbage collection''
time. High garbage collection overheads are generally an indication that
you are short of memory. The time excluding garbage collection is expected
to be tolerably consistent, but even then it can vary to 5 to 10 percent
even on a single computer. Such issues as the exact position and layout
of files on the disc or other background activity that the computer runs at
the same time have such effects.

\section{Removal}
For Unix you can remove REDUCE by just removing the \verb+r37+ directory and
all its contents, and any symbolic links you made into it. For Windows
de-installation is via the ``add/remove~programs'' item on the control panel
as usual.

\section{Applying patches}
From time to time minor updates and corrections to REDUCE will be published:
these can be located via the REDUCE home page, which is
\verb+http://www.rrz.uni-koeln.de/REDUCE/+.  The corrections will be
present as a downloadable file called \verb+patches.red+. Comments in this
file should explain what changes are being made, but the bulk of the
material there is not intended for the casual reader.  To install the
patches you should first locate the patches file in the REDUCE file
structure. It should be \verb+r37/packages/support/patches.red+. Make
a safe backup copy of this file, and replace it with the new copy
that you have downloaded. Read the comments in the new patches file to
see what has changed. Note that whenever you install new patches you can
expect some changes in the REDUCE test logs, at least in formatting and
layout.  Now go
\begin{verbatim}
   cd r37/lisp/csl/linux  (or win32, or whatever)
   # back up r37.img for safety here, please
   sh ../util/patchup     (or ..\util\patchup)
\end{verbatim}
REDUCE should run for a short time and update \verb+r37.img+ with the
new patches. When you next launch REDUCE its startup banner should
reflect the new date associated with the version of the patches file
you have just incorporated.  Note that there is no way to remove a set of
patches short of re-building the whole of REDUCE from source: just
putting back an older \verb+patches.red+ file and re-applying that is not
guaranteed to undo all effects of an intermediate patch.  So keep a copy of
your original \verb+r37.img+ so you can re-instate that if you have any
trouble. That should be the only file changed by the \verb+patchup+ job.

\section{Re-building from the REDUCE sources}
If you have a copy of the Professional Version of REDUCE it comes complete
with all source files. Users of the Personal system do not have the files
needed to do this and can ignore the rest of this document.
You may wish to recompile either just one REDUCE
module or the whole system. This will mainly be the case if you are
developing new packages for REDUCE.  The recipe is
\begin{verbatim}
   cd r37/lisp/csl/linux  (or win32, or whatever)
   ../util/full37
\end{verbatim}
\noindent This runs for a two or three minutes on the reference system,
and generates a log file in \verb+log/full37.log+. The version of
\verb+r37.img+ that it re-creates should have all current patches
installed.

To re-compile just a single REDUCE package, for instance \verb.groebner.,
use the sequence
\begin{verbatim}
   cd r37/lisp/csl/linux  (or win32, or whatever)
   ../util/package groebner
\end{verbatim}


\section{Re-compiling the CSL Lisp system}
REDUCE is built on top of a Lisp system: in this place that Lisp is called
CSL and all its sources are included in the Professional Version. When
distributed in this way the CSL Lisp system is intended for use just in
support of REDUCE so detailed information about its capabilities and
support for it (apart for as a component of REDUCE) is not provided.
The core parts of CSL are coded in the C language. This makes it
(fairly) easy to move CSL and hence REDUCE to new computer architectures
provided they have reliable C compilers and can give enough of an illusion
that they support 32-bit code.  Image files such as \verb+r37.img+ can be
created using a CSL that runs on one computer architecture and re-loaded
on another. Thus to mount REDUCE on another type of computer you just need
to compile this C code.  Create a new directory, calling it
\verb+r37/lisp/csl/my-machine+ and select it as the current directory.
Select one of the ready-made version of  \verb+Makefile+ from
\verb+../util+ and copy it of link it into this new directory. If your
local configuration is not exactly the same as one of the ones already catered
for you will need to edit the \verb+Makefile+ by hand. If you are attempting
this sort of re-compilation it is assumed that you already know enough to
sort out the details of that for yourself. In particular you may well find
that your C compiler needs some special flags setting or (even more probable)
that custom directives are needed to get all relevant libraries scanned.

For use on Windows you will find \verb+Makefile.w32+ is set up to use
the Watcom C compiler (tested using version 11), and \verb+Makefile.vc+
to use Microsoft's Visual C++ (tested using version 5).  \verb+Makefile.gcc+
is a good starting place for a generic Unix port using the widely
available free GNU C compiler.

It should then be the case that just saying \verb+make r37+ (or
\verb+make r37.exe+ in the Windows case) should build the relevant
executable.  There is no guarantee that the source code will compile
either first time or correctly on any system other than the
one you specified when you originally obtained it, and porting to new
architectures can call for changes in various system-dependent parts of
the code. Thus this level of re-compilation is intended to provide flexibility
for the expert who can cope with such issues for themselves rather than being
a fully-supported and guaranteed recipe for use by novices. In particular
in the past newer versions of C compilers have sometimes been incompatible
with older versions from the same vendor and Codemist does not guarantee
to keep the CSL C sources updated to cope with all such possible oddities.

As previously mentioned, when you have a new \verb+r37+ executable you can
copy and existing \verb+r37.img+ to the directory it lives in (REDUCE looks
for this image file in the directory it finds its executable in) and
test the system.

Of course even though you have the capability to re-compile all of REDUCE
in this way that does not mean you have permission to use it beyond the
terms of your license, and in particular you may not distribute versions
of the software that you have compiled for existing or new machines.

\section{Creating new profile information}
Parts of the REDUCE source code are translated into C and incorporated as
part of the CSL Lisp system. Doing this helps with performance. After
major changes to the REDUCE sources it may be useful to be able to
review which parts of REDUCE deserve this optimisation and to re-create the
C.  Doing this is a fairly costly business and only a very few users are
liable to want to attempt it.  The recipe (shows as done on Windows this
time) is
\begin{verbatim}
   cd r37\lisp\csl\win32
   make slowr37.exe
   ..\util\boot37
\end{verbatim}
\noindent This makes a special version of REDUCE that has full functionality
but which is much larger and slower than the final version. It does not
have any parts of it compiled into C.
\begin{verbatim}
   ..\util\profile
\end{verbatim}
\noindent The job that profiles REDUCE runs for over an hour on the
reference machine. It creates a file \verb+profile.dat+ in the current
directory. This file lists the most heavily used functions as revealed by
running all the REDUCE test scripts.  You then need to copy this file to
the place where the standard copy of it lives and use it to guide
selective compilation into C:
\begin{verbatim}
   copy profile.dat ..\csl-c
   ..\util\c-code37
\end{verbatim}
\noindent Finally you must re-compile \verb+r37.exe+ and re-build \verb+r37.img+ to
match. If you do one of these but not the other the system can be in
an incoherent state and may crash arbitrarily:
\begin{verbatim}
   make r37.exe
   ..\util\full37
\end{verbatim}
\noindent After such a major re-build it would be prudent to run all the
tests again:
\begin{verbatim}
   del ..\..\..\log\*.rlg
   ..\util\testall
   ..\util\checkall
\end{verbatim}
\noindent and inspect \verb+checkall.log+ to see that all is well.
\end{document}


