% The REDUCE User's Manual --- LaTeX version.
% Codemist Version with additional material in the same volume
% To create this manual, the following steps are recommended:
% latex r38
% bibtex r38
% latex r38
% latex r38
% makeindex r38
% latex r38
% dvipdfm r38

%% Does not contain
%% bibl.tex

\documentclass[11pt,letterpaper]{book}

\usepackage{makeidx}
% \usepackage{times}
\usepackage[dvipdfm]{graphicx}
\usepackage[dvipdfm]{hyperref}
\hyphenation{unique}
\hyphenation{effect}
\hyphenation{Stand-ard}
\hyphenation{libr-ary}
\hyphenation{direct-ory}
\hyphenation{state-ment}
\hyphenation{argu-ment}
\hyphenation{oper-ators}
\hyphenation{symb-olic}
\hyphenation{needs}
\hyphenation{GVARSLAST}
\hyphenation{ODE-SOLVE}
\hyphenation{hyper-geometric}
\hyphenation{equat-ion}
\hyphenation{equat-ions}
\hyphenation{OFF}
\hyphenation{Opt-ions}
\hyphenation{execu-tion}
\hyphenation{poly-nom-ials}
\hyphenation{func-t-ions}
\hyphenation{Inte-grals}
\hyphenation{Stutt-gart}

% More space in TOC requires this in book.sty
%\def\l@section{\@dottedtocline{1}{1.5em}{2.8em}}
%\def\l@subsection{\@dottedtocline{2}{4.3em}{3.2em}}
%\def\l@subsubsection{\@dottedtocline{3}{7.5em}{4.2em}}
%\def\l@paragraph{\@dottedtocline{4}{10.5em}{5em}}
%\def\l@subparagraph{\@dottedtocline{5}{12.5em}{6em}}

\setlength{\parindent}{0pt}

\setlength{\parskip}{6pt}

\setlength{\hfuzz}{5pt}  % don't complain about tiny overfull boxes
\setlength{\vfuzz}{1pt}

\renewcommand{\sloppy}{\tolerance=9999\relax%}
                       \setlength{\emergencystretch}{0.2\hsize}}

\tolerance=1000

\raggedbottom

\newlength{\reduceboxwidth}
\setlength{\reduceboxwidth}{4in}

\newlength{\redboxwidth}
\setlength{\redboxwidth}{3.5in}

\newlength{\rboxwidth}
\setlength{\rboxwidth}{2.6in}

\newcommand{\REDUCE}{REDUCE}
\newcommand{\RLISP}{RLISP}
\newcommand{\underscore}{\_}
\newcommand{\ttindex}[1]{{\renewcommand{\_}{\protect\underscore}%
                          \index{#1@{\tt #1}}}}
\newcommand{\COMPATNOTE}{{\em Compatibility Note:\ }}
% \meta{...} is an alternative sentential form in descriptions using \it.
\newcommand{\meta}[1]{\mbox{$\langle$\it#1\/$\rangle$}}

% Will print out a heading in bold, and then indent the following text.
\def\indented{\list{}{
 \itemindent\listparindent
 \rightmargin\leftmargin}\item[]}
\let\endindented=\endlist
\newenvironment{describe}[1]{\par{\bf #1}\begin{indented}}{\end{indented}}

% Close up default vertical spacings:
\setlength{\topsep}{0.5\baselineskip}  % above and below environments
\setlength{\itemsep}{\topsep}
\setlength{\abovedisplayskip}{\topsep}  % for "long" equations
\setlength{\belowdisplayskip}{\topsep}

\newcommand{\key}[1]{\fbox{\sf #1}}
\newcommand{\extendedmanual}[1]{#1}

\pagestyle{empty}

\makeindex

\begin{document}
\pagestyle{empty}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BeginCodemist

\vspace*{2.0in}
\begin{center}

{\Huge\bf {\REDUCE}} \\ [0.2cm]
{\LARGE\bf User's and \\
  Contributed Packages Manual\vspace{0.4cm} \\
  Version 3.8}

\vspace{0.5in}\large\bf

Anthony C.\ Hearn \\
Santa Monica, CA \\
and Codemist Ltd.

\vspace{0.1in}

\bf Email: reduce@rand.org

\vspace{0.5in}

\large\bf July 2003

\end{center}

\newpage
\vspace*{3.0in}
\noindent Copyright \copyright 2004 Anthony C. Hearn.  All rights reserved. \\
\mbox{}\\
%
\noindent Registered system holders may reproduce all or any part of this
publication for internal purposes, provided that the source of the
material is clearly acknowledged, and the copyright notice is retained.

\newpage

\pagestyle{headings}

\centerline{\bf \large Preface}

This volume has been prepared by Codemist Ltd. from the {\LaTeX}
documentation sources distributed with {\REDUCE} 3.8.  It incorporates
the User's Manual, and documentation for all the User Contributed
Packages as a second Part.  A common index and table of contents has been
prepared.  We hope that this single volume will be more convenient for
{\REDUCE} users than having two unrelated documents.  Particularly in
Part 2 the text of the authors has been extensively edited and
modified and so the responsibility for any errors rests with us.

Parts I and III were written by Anthony C. Hearn.  Part II is based on
texts by:\\
Werner Antweiler,
Victor Adamchik,
Joachim Apel,
Alan Barnes,
Andreas Bernig,
Yu.~A.~Blinkov,
Russell Bradford,
Chris Cannam,
Hubert Caprasse,
C.~{Dicrescenzo},
Alain Dresse,
Ladislav Drska,
James W.~Eastwood,
John Fitch,
Kerry Gaskell,
Barbara L.~Gates,
Karin Gatermann,
Hans-Gert Gr\"abe,
David Harper,
David {H}artley,
Anthony C.~Hearn,
J.~A.~van Hulzen,
V.~Ilyin,
Stanley L.~Kameny,
Fujio Kako,
C.~Kazasov,
Wolfram Koepf,
A.~Kryukov,
Richard Liska,
Kevin McIsaac,
Malcolm A.~H.~MacCallum,
Herbert Melenk,
H.~M.~M\"oller,
Winfried Neun,
Julian Padget,
Matt Rebbeck,
F.~Richard-Jung,
A.~Rodionov,
Carsten and Franziska Sch\"obel,
{Rainer} Sch\"opf,
Stephen Scowcroft,
Eberhard Schr\"{u}fer,
Fritz Schwarz,
M.~Spiridonova,
A.~Taranov,
Lisa Temme,
Walter Tietze,
V.~Tomov,
E.~Tournier,
Philip A.~Tuckey,
G.~\"{U}\c{c}oluk,
Mathias Warns,
Thomas Wolf,
Francis J.~Wright
and
A.~Yu.~Zharkov.

\noindent
\rightline{February 2004} \\
Codemist Ltd \\
``Alta'', Horsecombe Vale \\
Combe Down \\
Bath, England
\newpage
\tableofcontents
\part{{\REDUCE} User's Manual}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% EndCodemist

