\chapter{REDLOG: Logic System}
\label{REDLOG}
\typeout{{REDLOG: Logic System}}


{\footnotesize
\begin{center}
Andreas Dolzmann \\
Thomas Sturm \\
University of Passau, Germany \\
e-mail: dolzmann@uni-passau.de, sturm@uni-passau.de
\end{center}
}
\ttindex{REDLOG}

\section{Introduction}
This package extends \REDUCE\ to a computer logic system implementing
symbolic algorithms on first-order formulas wrt.~temporarily fixed
first-order languages and theories.

\subsection{Contexts}
REDLOG is designed for working with several languages and theories in
the sense of first-order logic. Both a language and a theory make up a
context. There are the following contexts available:
\begin{description}
\item[\textsc{OFSF}]
\textsc{OF} stands for \emph{ordered fields}, which is a little imprecise.
The quantifier elimination actually requires the more restricted class
of \emph{real closed fields}, while most of the tool-like algorithms
are generally correct for ordered fields. One usually has in mind real
numbers with ordering when using \textsc{OFSF}.

\item[\textsc{DVFSF}]
\emph{Discretely valued fields}. This is for computing with formulas
over classes of $p$-adic valued extension fields of the rationals,
usually the fields of $p$-adic numbers for some prime $p$.

\item[\textsc{ACFSF}]
\emph{Algebraically closed fields} such as the complex numbers.
\end{description}

\subsection{Overview}
REDLOG origins from the implementation of quantifier elimination
procedures. Successfully applying such methods to both academic and
real-world problems, the authors have developed over the time a large
set of formula-manipulating tools, many of which are meanwhile
interesting in their own right:

\begin{itemize}
\item
Numerous tools for comfortably inputing, decomposing, and analyzing
formulas.
\item
Several techniques for the \emph{simplification} of formulas.
\item
Various \emph{normal form computations}. The
\emph{\textsc{CNF}/\textsc{DNF}} computation includes both Boolean and
algebraic simplification strategies. The \emph{prenex normal form}
computation minimizes the number of quantifier changes.
\item
\emph{Quantifier elimination} computes quantifier-free equivalents for
given first-order formulas. For \textsc{OFSF} and \textsc{DVFSF} the
formulas have to obey certain degree restrictions.
\item 
The context \textsc{OFSF} allows a variant of quantifier elimination
called \emph{generic quantifier elimination}: There are certain
non-degeneracy assumptions made on the parameters, which considerably
speeds up the elimination.
\item
The contexts \textsc{OFSF} and \textsc{DVFSF} provide variants of
(generic) quantifier elimination that additionally compute
\emph{answers} such as satisfying sample points for existentially
quantified formulas.
\item
\textsc{OFSF}
includes linear \emph{optimization} techniques based on quantifier
elimination.
\end{itemize}

To avoid ambiguities with other packages, all \textsc{REDLOG} functions and
switches are prefixed by ``\texttt{RL}''.

The package is loaded by typing: \qquad {\tt load\_package redlog;} \\

It is recommended to read the documentation which comes with this
package. This manual chapter gives an overview on the features of
\textsc{REDLOG}, which is by no means complete.


\section{Context Selection}
The context to be used has to be selected explicitly. One way 
to do this is using the command \f{RLSET}\ttindex{RLSET}. As argument it takes one 
of the
valid choices \f{ACFSF}\ttindex{ACFSF} (algebraically closed fields
standard form),
\f{OFSF}\ttindex{OFSF} (ordered fields standard form), and 
\f{DVFSF}\ttindex{DVFSF}
(discretely valued fields standard form). By default, \f{DVFSF}\ttindex{DVFSF} 
computes
uniformly over the class of all $p$-adic valued fields. For the sake
of efficiency, this can be restricted by means of an extra 
\f{RLSET}\ttindex{RLSET} argument.
\f{RLSET}\ttindex{RLSET} returns the old setting as a list.

\section{Format and Handling of Formulas}

\subsection{First-order Operators}
REDLOG knows the following operators for constructing Boolean
combinations and quantifications of atomic formulas:

\begin{center}
\begin{tabular}{llll}
\f{NOT}\ttindex{NOT}: Unary & 
\f{AND}\ttindex{AND}: N-ary Infix &
\f{OR}\ttindex{OR}: N-ary Infix  &
\f{IMPL}\ttindex{IMPL}: Binary Infix \\
\f{REPL}\ttindex{REPL}: Binary Infix  & 
\f{EQUIV}\ttindex{EQUIV}: Binary Infix &
\f{EX}\ttindex{EX}: Binary \\
\f{ALL}\ttindex{ALL}: Binary &
\f{TRUE}\ttindex{TRUE}: Variable &
\f{FALSE}\ttindex{FALSE}: Variable & 
\end{tabular}
\end{center}

The \f{EX} and the \f{ALL} operators are the quantifiers. Their first
argument is the quantified variable, the second one a matrix formula.

There are operators \f{MKAND}\ttindex{MKAND} and
\f{MKOR}\ttindex{MKOR} for the construction of large systematic
conjunctions/disjunctions via for loops available. They are used in
the style of \f{SUM} and \f{COLLECT}.

\vspace{0.5cm}
{\bf Example:}
\begin{verbatim}
1: load_package redlog;

2: rlset ofsf;

{}

3: g := for i:=1:3 mkand

          for j:=1:3 mkor

            if j<>i then mkid(x,i) + mkid(x,j)=0;

true and (false or false or x1 + x2 = 0 or x1 + x3 = 0)

 and (false or x1 + x2 = 0 or false or x2 + x3 = 0)

 and (false or x1 + x3 = 0 or x2 + x3 = 0 or false)
\end{verbatim}

\subsection{OFSF Operators}
The \f{OFSF}\ttindex{OFSF} context implements {\it ordered fields}
over the language of {\it ordered rings}. There are the following
binary operators available:

\begin{center}
\begin{tabular}{llllllll}
\f{EQUAL}\ttindex{EQUAL} &
\f{NEQ}\ttindex{NEQ} &
\f{LEQ}\ttindex{LEQ} &
\f{GEQ}\ttindex{GEQ} &
\f{LESSP}\ttindex{LESSP} & 
\f{GREATERP}\ttindex{GREATERP}
\end{tabular}
\end{center}
They can also be written as \f{=}, \f{<>}, \f{<=}, \f{>=}, \f{<}, and
\f{>}.
For {\sc OFSF}
there is specified that all right hand sides must be zero. Non-zero right
hand sides are immediately subtracted. 

\subsection{DVFSF Operators}\ttindex{DVFSF}
Discretely valued fields are implemented as a one-sorted language
using in addition to \f{=} and \f{<>} the
binary operators \f{|}, \f{||}, \f{\~{}}, and \f{/\~{}}, which encode
$\leq$, $<$, $=$, and $\neq$ in the
value group, respectively.
\begin{center}
\begin{tabular}{llllll}
\f{EQUAL}\ttindex{EQUAL} &
\f{NEQ}\ttindex{NEQ} &
\f{DIV}\ttindex{DIV} &
\f{SDIV}\ttindex{SDIV} &
\f{ASSOC}\ttindex{ASSOC} & 
\f{NASSOC}\ttindex{NASSOC} \\
\end{tabular}
\end{center}

\subsection{ACFSF Operators}\ttindex{ACFSF}
For algebraically closed fields there are only equations and
inequalities allowed:

\begin{center}
\begin{tabular}{ll}
\f{EQUAL}\ttindex{EQUAL} &
\f{NEQ}\ttindex{NEQ}
\end{tabular}
\end{center}

As in \textsc{OFSF}, they can be conveniently written as \f{=} and
\f{<>}, respectively. All right hand sides are zero.

\subsection{Extended Built-in Commands}
The operators 
\f{SUB}\ttindex{SUB},
\f{PART}\ttindex{PART},
and \f{LENGTH}\ttindex{LENGTH} work on formulas in a reasonable way.

\subsection{Global Switches}
The switch \f{RLSIMPL}\ttindex{RLSIMPL} causes the function
\f{RLSIMPL} to be automatically applied at the expression evaluation stage.

The switch \f{RLREALTIME}\ttindex{RLREALTIME} protocols the wall clock
time needed for {\sc REDLOG} commands in seconds.

The switch \f{RLVERBOSE}\ttindex{RLVERBOSE} toggles verbosity output
with some {\sc REDLOG} procedures.

\section{Simplification}
{\sc REDLOG} knows three types of simplifiers to reduce the size of a
given first-order formula: the standard simplifier, tableau
simplifiers, and Gr\"obner simplifiers.

\subsection{Standard Simplifier}
The standard simplifier \f{RLSIMPL}\ttindex{RLSIMPL} returns a
simplified equivalent of its argument formula. It is much faster
though less powerful than the other simplifiers.

As an optional argument there can be a \emph{theory} passed. This is a
list of atomic formulas assumed to hold. Simplification is then
performed on the basis of these assumptions.

\vspace{0.5cm}
{\bf Example:}
\begin{verbatim}
4: rlsimpl g;

(x1 + x2 = 0 or x1 + x3 = 0) and (x1 + x2 = 0 or x2 + x3 = 0)

 and (x1 + x3 = 0 or x2 + x3 = 0)
\end{verbatim}

\subsection{Tableau Simplifier}
The standard simplifier preserves the basic Boolean structure of a formula. The
tableau methods, in contrast, provide a technique for changing the Boolean 
structure of a formula by constructing case distinctions.

The function \f{RLATAB}\ttindex{RLATAB} automatically finds a suitable
case distinction. Based on \f{RLATAB}, the function
\f{RLITAB}\ttindex{RLITAB} iterates this process until no further
simplification can be detected. There is a more fundamental entry
point \f{RLTAB}\ttindex{RLTAB} for manually entering case
distinctions.

\subsection{Gr\"obner Simplifier}
The Gr\"obner simplifier considers algebraic simplification rules
between the atomic formulas of the input formula. The usual procedure
called for Gr\"obner simplification is \f{RLGSN}\ttindex{RLGSN}.
Similar to the standard simplifier, there is an optional theory
argument.

\begin{samepage}
\vspace{0.5cm}
{\bf Example:}
\begin{verbatim}
5: rlgsn(x*y+1<>0 or y*z+1<>0 or x-z=0);

true
\end{verbatim}
\end{samepage}

\section{Normal Forms}

\subsection{Boolean Normal Forms}
\f{RLCNF}\ttindex{RLCNF} and \f{RLDNF}\ttindex{RLDNF} compute conjunctive
resp.~disjunctive normal forms of their formula arguments. Subsumption
and cut strategies are applied to decrease the number of clauses.

\subsection{Miscellaneous Normal Forms}
\f{RLNNF}\ttindex{RLNNF} computes a 
negation normal form. This is an {\tt and}-\texttt{or}-combination of
atomic formulas.

\f{RLPNF}\ttindex{RLPNF} computes a prenex normal form of its
argument. That is, all quantifiers are moved outside such that they
form a block in front of a quantifier-free matrix formula.

\section{Quantifier Elimination and Variants}
Quantifier elimination computes quantifier-free equivalents for given
first-order formulas. For \textsc{OFSF} and \textsc{DVFSF}, REDLOG
uses a technique based on elimination set ideas. The \textsc{OFSF}
implementation is restricted to at most quadratic occurrences of the
quantified variables, but includes numerous heuristic strategies for
coping with higher degrees. The \textsc{DVFSF} implementation is
restricted to formulas that are linear in the quantified variables.
The \textsc{ACFSF} quantifier elimination is based on comprehensive
Gr\"obner basis computation; there are no degree restrictions for this
context

\subsection{Quantifier Elimination}
\f{RLQE}\ttindex{RLQE} performs quantifier elimination on its argument
formula. There is an optional theory argument in the style of
\f{RLSIMPL} supported.

\begin{samepage}
\vspace{0.5cm}
{\bf Example:}
\begin{verbatim}
6: rlqe(ex(x,a*x**2+b*x+c>0),{a<0});

         2
4*a*c - b  < 0
\end{verbatim}
\end{samepage}

For \textsc{OFSF} and \textsc{DVFSF} there is a variant
\f{RLQEA}\ttindex{RLQEA} available. It returns instead of a
quantifier-free equivalent, a list of condition-solution pairs
containing, e.g., satisfying sample points for outermost existential
quantifier blocks.

\begin{samepage}
\vspace{0.5cm}
{\bf Example:}
\begin{verbatim}
7: rlqea(ex(x,a*x**2+b*x+c>0),{a<0});

           2
{{4*a*c - b  < 0,

                            2
         - sqrt( - 4*a*c + b ) - 2*a*epsilon1 - b
  {x = -------------------------------------------}}}
                           2*a
\end{verbatim}
\end{samepage}

\subsection{Generic Quantifier Elimination}
\textsc{OFSF} allows generic quantifier elimination
\f{RLGQE}\ttindex{RLGQE}, which enlarges the theory by disequations,
i.e.~\f{<>}-atomic formulas, wherever this supports the quantifier
elimination. There is also generic quantifier elimination with answer
available: \f{RLGQEA}\ttindex{RLGQEA}.

\begin{samepage}
\vspace{0.5cm}
{\bf Example:}
\begin{verbatim}
8: rlgqe ex(x,a*x**2+b*x+c>0);

{{a <> 0},

          2
 4*a*c - b  < 0 or a >= 0}
\end{verbatim}
\end{samepage}

\subsection{Linear Optimization}
\f{RLOPT}\ttindex{RLOPT} uses quantifier elimination for linear
optimization. It takes as arguments a list of constraints and the
target function. The target function is minimized subject to the
constraints.
